---
title: "Ratings of Tree Species Environmental Suitability"
authors: "William H MacKenzie, Allen Banner, Del V. Meidinger, Hardy Griesbauer, Ray Coupe, Heather Klassen, Sari Saunders, Tom Braumandl, Bruce Rogers, Richard Kabzems"
date: "5/12/2024"
format:
  typst:
    toc: true
    toc-depth: 2
    toc-title: Contents
    section-numbering: 1.1.1
    columns: 1
    mainfont: Minion Pro
    fontsize: 10pt
editor: source
execute:
  echo: false
  error: false
  warning: false
  message: false
  fig.width: 6
  fig.height: 4
  fig.align: 'left'
  fig.cap: true
  fig.pos: H
  out.width: '100%'
  dev: pdf
  fig.ext: pdf
  cache: true
  fig.retina: 2
  dpi: 600
  fig.asp: 1.5
  fig.path: "./figures/"
---

```{r}
require(ggplot2)
require(tidyverse)
require(gt)
```

# Introduction

The Chief Forester’s Reference Guide (CFRG) applies a four-class suitability rating system (primary, secondary, tertiary, Unrated/unsuitable) to each tree species in every recognized site series that occurs in British Columbia. These suitability ratings are defined as ‘ecological suitability for a timber objective’, and based on an assessment of three factors: environmental suitability of sites, forest health considerations, and ability to supply timber for sawlog production. These ratings were used historically to inform the identification of preferred and acceptable species in the creation of stocking standards.

There is no consistent or documented approach to how a suitability rating is created from an assessment of thee different factors. The CFRG suitability ratings are generally higher for preferred timber species (e.g., Douglas-fir) and often under-rated for less desirable species (e.g., subalpine fir). Broad-leaved species, which do not produce sawlog timber, are not rated in a similar fashion. Additionally, the CFRG ratings are adjusted within a site series to always present at least one primary species as the “best available species” for sawlog production in a site series.

Thus, use of the CFRG species suitability ratings for creation of alternative (non-sawlog) management prescriptions or for climate change projections of tree species performance is suboptimal. Rather, these uses will benefit from a independent evaluation of each of 3 factors (environmental feasibility + reliability/forest health factors + management goal/intent). Environmental feasibility of tree species at a site series level is the primary factor in this assessment as it identifies which species are candidates for consideration in the management prescription as well as providing some information useful for evaluating reliability.

This document describes an approach for assessing the environmental feasibility of native tree species (aka: ‘species feasibilities’) in the conditions of a site series, and for the creation of a species by site series feasibility rating matrix for British Columbia and adjacent jurisdictions. This set of species feasibility ratings will then be available for assessing tree species for use in non-sawlog management prescriptions and for climate change modelling. Forest health considerations are to be reviewed separately by forest health specialists for inclusion in reforestation decisions. Management prescriptions to different management objectives can then be created from these ratings.

# Species Feasibility Ratings

**Environmental feasibility** ratings refer to the degree to which a tree species can successfully establish and develop under specific environmental conditions (e.g., site series).

**Environmental suitability** refers to the capacity of a site to support a tree species.

The species feasibility ratings are focused on the establishment phase of reforestation when trees are most vulnerable to environmental stresses, but these ratings should also reflect likely success through to maturation (assuming non-climate changed future). The primary rating criterion is an assessment of the probability of successful establishment and growth of a species across the entire environmental space of a site series. It is assumed that the species lists in the CFRG are an accurate list of the feasible species for each site series. However, the CFRG ratings reflect “best species” for saw log production or “best available species” for a site series. For this reassessment, species are initially assumed to be feasible across all examples of a site series (an E1 rating). A species may then be down-rated using several criteria indicating environmental constraints (e.g., CFRG footnotes indicating site-limits or topographic or geographic restrictions, observed 'decline' of species at/post maturity). Deciduous tree species, which have been less comprehensively rated in the CFRG, are given a similar and thorough rating treatment as conifers through all BGCs.

The environmental feasibility rankings represent a simplified response curve over the environmental space of the species (@fig-env-response).

```{r conceptual env response curve, echo=FALSE}
#| label: fig-env-response
#| fig-cap: "Conceptual relationship of Environmental Feasibility Ratings on a response curve"
#| fig-cap-location: bottom
#| warning: false
#| fig-align: "center"
# Generate data for the curve
x <- seq(-3, 3, length=400)
y <- exp(-x^2)

# Create a data frame
data <- data.frame(x, y)

# Plot the curve
# Define labels for each segment
labels_data <- data.frame(
  x = c(-2.625, -2, -1.5, -1, 0, 1, 1.5, 2, 2.625),
  y = rep(max(y) + 0.05, 9),
  label = c("E5", "E4", "E3", "E2", "E1", "E2", "E3", "E4", "E5")
)

# Create the plot with vertical lines and labels at specified positions
ggplot(data, aes(x, y)) +
  geom_line(color = "grey20", size = 1) +
  geom_vline(xintercept = c(-3, -2.25, -1.75, -1.25, -0.75, 0.75, 1.25, 1.75, 2.25, 3), linetype = "dashed", size=.75, color = "grey20") +
  geom_text(data = labels_data, aes(x = x, y = y, label = label), color = "black", fontface = "bold") +
  labs(x = "Environmental Gradient",
       y = "Environmental Feasibility") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.border = element_rect(color = "black", fill = NA))
```

## Feasibility Rating Definitions

-   **E1** – High feasibility: species well within its environmental tolerance range.
    -   The species is generally common and abundant in natural forests but may be of variable abundance where disturbance regime/stand age distribution is unfavorable (e.g., shade intolerant species in low disturbance landscapes).
    -   Widespread establishment/regeneration success on all site conditions representative of the site series. Expect continuous distribution at a landscape or stand level where the species has been established.
    -   Good growth rate and form. (*\>=50% of max SI for species?*)
    -   Typically no geographic or environmental limitation footnotes appear in the Reference Guide.
-   **E2** – Moderate feasibility: species occurring outside the core environmental tolerance range for the species.
    -   Species may be at the periphery of its climate range distribution or its edaphic tolerance.
    -   The species has variable abundance in natural forests and is often concentrated in select site types of the site series (e.g., warm aspects, higher elevations, elevated micro sites). Not all site conditions of the site series or normal post-disturbance stand conditions may be suitable for successful establishment/regeneration. Patchy or zoned distribution at a landscape and/or stand level are common.
    -   The species may have good growth and survival where successfully established on suitable sites. However, E2 rating may be used to reflect site series with broadly unsuitable site conditions across species (Spruce-horsetail wetland phases, xeric site series with shallow soils) or BGCs with broadly reduced growth potential (woodland subzones) where slower growth rate and possible poor form are common.
    -   Geographic, topographic, or other environmental limitation footnotes are commonly applied in the Reference Guide.
-   **E3** – Low feasibility: species occurring at the edges of the species climatic or site tolerance.
    -   Species likely to be adjacent to climates or site conditions where the species is unsuitable
    -   The species is generally infrequent to sporadic in natural forests and has significant environmental limitations in the site series. Only specific landscape positions or microsites in a site series are likely to have establishment/regeneration success, but subsequent good growth is possible; or
    -   E3 applied in BGCs with very limited growth potential (e.g. parkland subzones) or have harsh site conditions for trees (e.g. xeric sites with shallow discontinuous soils)
    -   Multiple limiting factors in the CFRG are common. Comments on the limited viability of the site series for forestry may reflect E3 ratings for all species in that site series.
    -   Very slow growth and/or deformed. Considerably restricted tree growth potential.
-   **E4** – Minimal feasibility. Species that occur at the threshold of environmental tolerance
    -   Scattered occurence in natural stands often not reaching main canopy (e.g Hw in the western SBSmc2)
    -   Off-site trials with persistent but low survival and poor growth
    -   in but not complete failureexpected to be feasible outside of the species current natural range based on modelling or (preferably) successful field trials. These species may receive E1-E3 feasibility ratings upon further review of available information: Griesbauer and MacKenzie: Type 1 off-site species criteria.
-   **E5** – No/minimal environmental feasibility based on species autecology. Environmental constraints generally preclude successful establishment of untended stands. Species not present in natural forests. F5 has been assigned only if the species was previously ranked as suitable for a site series in the Reference Guide and is reassessed as unfeasible. F5 rating is assumed where a species has a null rating in the tables.

## Feasibility Rating Process

Initial feasibility ratings were assigned by A. Banner and T. Braumandl for all regions using the following information resources.

In some cases, this reassessment has resulted in commercially valued species receiving lower feasibility ratings relative to the suitability rating in the reference guide. Species indicative of a BGC unit may also receive a lower rating in some cases (e.g., Fd as an E2 in the IDFdk4 where frost and cold temperature limitations are widespread). Site series near the limits of tree species feasibility (parkland climates, sites with very thin soil veneers) will often have no tree species with highest feasibility (E1), due to general constraints to trees in these environmental conditions. These differences are not a problem, and generally we do not recommend directly comparing the feasibility and suitability ratings (e.g., E1 vs primary) as they are comprised of different decision-making factors.

Subsequent review of this output led to some clarification of definitions below and modifications to initial ratings attributes. An ordinal random forest model was trained from these ratings with climate and site-level environmental variables. This model was then used to predict suitability ratings for all site series not previously assessed. This process was also used to identify cases where specific species ratings were misclassified by the random forest model and might reflect inconsistent ratings within the applied ratings and should be reassessed. Site series with extra-edatopic factors (cold-air site series, floodplains, etc.) have been incompletely flagged in the dataset. Predictions from the random-forest model do not include these factors. D. Meidinger subsequently reviewed his suitability ratings for the USA_BGC and made modifications to align with the rating definitions below. Review of ratings required by regional experts.

## Species Feasibility Workflow

Add decision workflow draft here.

{{< pagebreak >}}
# Appendix 1. Examples of Environmental Feasibility Ratings Decisions

Some decision/discussion examples for environmental feasibility ratings:

1.  Bl and Sx/Se in ESSF forests. Both generally receive an E1 rating even where Bl dominates the mature forest. Regeneration of either species is likely to be highly successful. However, where growth potential is lower reflecting generally limiting environmental conditions, such as in woodland and parkland environments, lower feasibility ratings are applied to all site series, see point 2 below.

2.  Increasingly cold and snowy conditions with elevation are generally stressful for tree species in the woodland and parkland environments leading to increased tree spacing and reduced vigour and form (esp. parkland). Species in these BGCs have generally been given max suitability values of E2 and E3 respectively.

3.  Broad-leaved trees were not rated in the original reference guide as these species have no sawlog value. In the original AB review an E1 was only given where a species was considered the dominant tree species in the mature stage (e.g., Middle bench Floodplains). Ratings of ‘E2’ were given where the species would thrive on most sites in appropriately disturbed sites and an **E3** was given where the species has limited establishment success following disturbance. These guiding principles have subsequently been modified to apply an **E1** rating where widespread long-term deciduous or mixed wood stands occur (e.g., aspen in SBSdk, SBPSmk, BWBSmw, etc.).

4.  **Cw establishment** in open conditions (e.g., clearcut) is not generally an issue in coastal or wetter interior environments and is given a rating of **E1**. However, in some ICH climates, heat and sun exposure may limit the species in the establishment phase – these likely also represent climates/sites where the species has less suitability and an **E2** or lower rating may be assigned. Similar factors affect **Fd** where planting under partial canopy is recommended/required (FN 27).

5.  Where a species’ previous reference guide rating was lower due mainly to perceived silvicultural/wood quality concerns (e.g., Bl, Cw, Hw), a higher rating was often assigned based on environmental feasibility. (e.g., Bl previously assigned a E2 in some ESSF subzone/variants but now assigned an **E1** for some site series (e.g., SBSmc2, Kamloops ESSF)).

6.  Where a species’ previous reference guide rating was high due mainly to high timber desirability (e.g., Fd), but portions of the site series range are not considered of high feasibility, a lower rating is often applied even though on the suitable portions of the site series the species may perform well.
{{< pagebreak >}}
# Appendix II. Chief Forester's Reference Guide Footnotes

{{< pagebreak >}}
## Appendix IIa. Environment Limiting Footnotes

-   **H** = Heat related limitation (may include aridity)
-   **A** = Aridity limitation
-   **C** = Cold limitation
-   **S** = Snow limitations
-   **F** = Frost limitations
-   **T** = Maritime/Continentality Limits
-   **W** = Overly wet site conditions
```{r}
# Data for the table
data <- data.frame(
  Revised_Footnote = c(13, 10, 12, 15, 27, 69, 46, 207, 24, 28, 14, 9, 16, 205, 22, 32, 34, 66, 41, 55, 1, 17, 18, 20, 44, 45),
  Revised_Footnote_Text = c(
    "most suitable at upper elevations", "most suitable on cool aspects", "most suitable on cold air drainage sites",
    "most suitable in the northern portion of the BGC unit", "partial canopy cover required for establishment",
    "recommended at upper elevations only when planted in the southern portion of the BGC unit", "most suitable to area north of the Dean Channel",
    "obstacle planting recommended", "most suitable in wetter portion of the BGC unit", "limited by moisture deficit",
    "most suitable at lower elevations", "most suitable on warm aspects", "most suitable in the southern portion of the BGC unit",
    "limited by cold temperatures", "most suitable in the southern Gardner Canal-Kitlope area", "limited by growing-season frosts",
    "risk of snow damage", "Recommended as preferred where risk of snow damage is low or risk of frost damage on spruce is high",
    "limited by poorly-drained soils", "acceptable on subxeric and submesic sites", "most suitable on elevated microsites",
    "most suitable in the western portion of the BGC unit", "most suitable in the eastern portion of the BGC unit",
    "not suitable on the outer coast", "suitable in areas with strong maritime influence", "suitable in areas with strong continental influence"
  ),
  Footnote_Type_Code = c("H", "H", "H", "H", "H", "H", "H", "H", "A", "A", "C", "C", "C", "C", "C", "F", "S", "S", "W", "W", "W", "Tc", "Tm", "Tc", "Tc", "Tm")
)

# Create the gt table
gt_table <- gt(data) %>%
  tab_header(
    title = "Site or Climatic LimitationsFootnotes"#,
    #subtitle = "Site or Climatic Limitations"
  ) %>%
  cols_label(
    Revised_Footnote = "Revised Footnote",
    Revised_Footnote_Text = "Revised Footnote Text",
    Footnote_Type_Code = "Footnote_Type_Code"
  ) %>%
  fmt_markdown(columns = everything()) %>%
  tab_options(
    table.font.size = px(9),
    data_row.padding = px(2),
    column_labels.padding = px(2),
    table.width = pct(100)
  ) %>%
  cols_width(
    Revised_Footnote ~ px(70),
    Revised_Footnote_Text ~ px(300),
    Footnote_Type_Code ~ px(70
                            )
  )


# Print the gt table
gt_table
```
{{< pagebreak >}}
## Appendix IIb. Site or Soils Limiting Footnotes 

```{r}
# Data for the table
data2 <- data.frame(
  Revised_Footnote = c(60, 61, 19, 21),
  Revised_Footnote_Text = c(
    "Acceptable only in the Squamish District", "Acceptable only in the Squamish District on cold air drainage sites",
    "not suitable on Haida Gwaii", "suitable only on the mainland"
  ),
  NEW_Environmental_Limitation_Footnote = c("G", "G", "G", "G")
)

# Create the gt table
gt_table2 <- gt(data2) %>%
  tab_header(
    title = "Chief Forester Reference Guide Footnotes",
    subtitle = "Biogeographic Range Limitations"
  ) %>%
  cols_label(
    Revised_Footnote = "Revised Footnote",
    Revised_Footnote_Text = "Revised Footnote Text",
    NEW_Environmental_Limitation_Footnote = "NEW Environmental Limitation Footnote"
  ) %>%
  fmt_markdown(columns = everything()) %>%
  tab_options(
    table.font.size = px(9),
    data_row.padding = px(2),
    column_labels.padding = px(2),
    table.width = pct(100)
  ) %>%
  cols_width(
    Revised_Footnote ~ px(100),
    Revised_Footnote_Text ~ px(300),
    NEW_Environmental_Limitation_Footnote ~ px(150)
  )

# Print the gt table
gt_table2
```

## Appendix IIb. Site or Soils Limiting Footnotes 

```{r}
# Data for the table
data3 <- data.frame(
  Revised_Footnote = c(2, 3, 4, 6, 7, 8, 11, 39, 42, 52),
  Revised_Footnote_Text = c(
    "most suitable on thick forest floors", "most suitable on coarse-textured soils", "most suitable on medium-textured soils",
    "most suitable on nutrient-very poor sites", "most suitable on nutrient-medium sites", "most suitable on steep slopes",
    "most suitable on crest slope positions", "avoid exposed and windy sites", "most suitable on sites with a fresh soil moisture regime",
    "most suitable on sheltered microsites with deep soils"
  ),
  Environmental_Limitation = c("?", "?", "?", "?", "?", "?", "?", "?", "?", "?")
)

# Create the gt table
gt_table3 <- gt(data3) %>%
  tab_header(
    title = "Chief Forester Reference Guide Footnotes",
    subtitle = "Site or Soils Preferences"
  ) %>%
  cols_label(
    Revised_Footnote = "Revised Footnote",
    Revised_Footnote_Text = "Revised Footnote Text",
    Environmental_Limitation = "Environmental Limitation"
  ) %>%
  fmt_markdown(columns = everything()) %>%
  tab_options(
    table.font.size = px(9),
    data_row.padding = px(2),
    column_labels.padding = px(2),
    table.width = pct(100)
  ) %>%
  cols_width(
    Revised_Footnote ~ px(100),
    Revised_Footnote_Text ~ px(300),
    Environmental_Limitation ~ px(150)
  )

# Print the gt table
gt_table3
```


